---
title: "COMP4028 – Data Analysis using R Programming"
subtitle: "Food Allergy Analysis – Zenodo Dataset"
author: "Group 6"
date: "28 March 2026`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Introduction

This report performs a comprehensive data analysis on the **Food Allergy Analysis Dataset** sourced from [Zenodo](https://zenodo.org/). The dataset contains de-identified patient records tracking food allergy diagnoses, atopic conditions, and demographic information for over 333,000 subjects. The analysis is performed using R Programming and covers data exploration, manipulation, statistical analysis, and visualization.

---

# Load Required Libraries

```{r libraries}
library(dplyr)
library(ggplot2)
library(tidyr)
library(reshape2)
```

---

# Load the Dataset

```{r load-data}
# Read the gzipped CSV file directly
df <- read.csv("C:/Users/kvnsc/Downloads/food-allergy-analysis-Zenodo.csv", stringsAsFactors = FALSE)
```

---

# 1. Print the Structure of the Dataset

```{r structure}
str(df)
```

---

# 2. List the Variables in the Dataset

```{r variables}
names(df)
```

---

# 3. Print the Top 15 Rows

```{r head}
head(df, 15)
```

---

# 4. User-Defined Function

A custom function is defined to calculate the **duration** of a condition (e.g. asthma) given a start and end age column. It returns the duration in years, or `NA` if either value is missing.

```{r user-function}
# User-defined function: calculate duration between two age columns
calculate_duration <- function(start_age, end_age) {
  duration <- ifelse(!is.na(start_age) & !is.na(end_age),
                     round(end_age - start_age, 2),
                     NA)
  return(duration)
}

# Apply to asthma start and end ages
df$ASTHMA_DURATION <- calculate_duration(df$ASTHMA_START, df$ASTHMA_END)

# Preview results
head(df[!is.na(df$ASTHMA_DURATION), c("SUBJECT_ID", "ASTHMA_START", "ASTHMA_END", "ASTHMA_DURATION")], 10)
```

---

# 5. Filter Rows Based on Logical Criteria

Filter to keep only patients who:
- Are part of the **Atopic March Cohort** (`ATOPIC_MARCH_COHORT == TRUE`)
- Have a **recorded peanut allergy start** age

```{r filter}
filtered_df <- df %>%
  filter(ATOPIC_MARCH_COHORT == TRUE & !is.na(PEANUT_ALG_START))

cat("Rows after filtering:", nrow(filtered_df), "\n")
head(filtered_df[, c("SUBJECT_ID", "GENDER_FACTOR", "RACE_FACTOR",
                     "PEANUT_ALG_START", "ATOPIC_MARCH_COHORT")], 10)
```

---

# 6. Identify Dependent & Independent Variables and Reshape

**Independent variables:** `AGE_START_YEARS`, `GENDER_FACTOR`, `RACE_FACTOR`, `PAYER_FACTOR`  
**Dependent variable:** `NUM_ASTHMARX` (number of asthma prescriptions — outcome of interest)

A new data frame is created by joining these selected variables.

```{r reshape}
# Select independent and dependent variables
independent_vars <- df %>%
  select(SUBJECT_ID, AGE_START_YEARS, GENDER_FACTOR, RACE_FACTOR, PAYER_FACTOR)

dependent_var <- df %>%
  select(SUBJECT_ID, NUM_ASTHMARX)

# Join into a new data frame
analysis_df <- inner_join(independent_vars, dependent_var, by = "SUBJECT_ID")

cat("Dimensions of reshaped data frame:", dim(analysis_df), "\n")
head(analysis_df, 10)
```

---

# 7. Remove Missing Values

```{r remove-na}
cat("Rows before removing NAs:", nrow(analysis_df), "\n")

analysis_df_clean <- na.omit(analysis_df)

cat("Rows after removing NAs:", nrow(analysis_df_clean), "\n")
cat("Rows removed:", nrow(analysis_df) - nrow(analysis_df_clean), "\n")
```

---

# 8. Identify and Remove Duplicated Data

```{r duplicates}
cat("Number of duplicate rows:", sum(duplicated(df)), "\n")

# Remove duplicates
df_no_dupes <- df[!duplicated(df), ]

cat("Rows before deduplication:", nrow(df), "\n")
cat("Rows after deduplication:", nrow(df_no_dupes), "\n")
```

---

# 9. Reorder Rows in Descending Order

Reorder the cleaned analysis data frame by `NUM_ASTHMARX` in descending order (patients with the most asthma prescriptions first).

```{r reorder}
analysis_df_sorted <- analysis_df_clean %>%
  arrange(desc(NUM_ASTHMARX), desc(AGE_START_YEARS))

head(analysis_df_sorted, 10)
```

---

# 10. Rename Column Names

```{r rename}
analysis_df_renamed <- analysis_df_clean %>%
  rename(
    Patient_ID        = SUBJECT_ID,
    Age_At_Start      = AGE_START_YEARS,
    Gender            = GENDER_FACTOR,
    Race              = RACE_FACTOR,
    Insurance_Type    = PAYER_FACTOR,
    Asthma_Rx_Count   = NUM_ASTHMARX
  )

names(analysis_df_renamed)
head(analysis_df_renamed, 5)
```

---

# 11. Add New Variables Using a Mathematical Function

Add two new computed columns:
- `Asthma_Rx_Double`: number of asthma prescriptions multiplied by 2
- `Age_At_Start_Months`: age at start converted from years to months

```{r new-variables}
analysis_df_renamed <- analysis_df_renamed %>%
  mutate(
    Asthma_Rx_Double    = Asthma_Rx_Count * 2,
    Age_At_Start_Months = round(Age_At_Start * 12, 1)
  )

head(analysis_df_renamed[, c("Patient_ID", "Age_At_Start", "Age_At_Start_Months",
                              "Asthma_Rx_Count", "Asthma_Rx_Double")], 10)
```

---

# 12. Create a Training Set Using Random Number Generator

```{r training-set}
# Set seed for reproducibility
set.seed(42)

# Generate a random 70% training set
train_index <- sample(1:nrow(analysis_df_renamed),
                      size = floor(0.70 * nrow(analysis_df_renamed)),
                      replace = FALSE)

train_set <- analysis_df_renamed[train_index, ]
test_set  <- analysis_df_renamed[-train_index, ]

cat("Total rows:", nrow(analysis_df_renamed), "\n")
cat("Training set rows (70%):", nrow(train_set), "\n")
cat("Test set rows (30%):", nrow(test_set), "\n")

head(train_set, 5)
```

---

# 13. Print Summary Statistics

```{r summary}
summary(analysis_df_renamed)
```

---

# 14. Statistical Functions on Numerical Variables

Using `AGE_START_YEARS` (age at which the patient entered the study) and `NUM_ASTHMARX` (number of asthma prescriptions).

```{r stats}
# --- AGE_START_YEARS ---
age_values <- analysis_df_renamed$Age_At_Start

cat("=== Age At Start (Years) ===\n")
cat("Mean:  ", round(mean(age_values), 4), "\n")
cat("Median:", round(median(age_values), 4), "\n")
cat("Range: ", round(range(age_values), 4), "\n")

# Mode (custom function — R has no built-in mode for continuous data)
get_mode <- function(x) {
  x_clean <- x[!is.na(x)]
  unique_vals <- unique(x_clean)
  unique_vals[which.max(tabulate(match(x_clean, unique_vals)))]
}
cat("Mode:  ", round(get_mode(age_values), 4), "\n")

# --- NUM_ASTHMARX ---
rx_values <- analysis_df_renamed$Asthma_Rx_Count

cat("\n=== Number of Asthma Prescriptions ===\n")
cat("Mean:  ", round(mean(rx_values), 4), "\n")
cat("Median:", round(median(rx_values), 4), "\n")
cat("Range: ", range(rx_values), "\n")
cat("Mode:  ", get_mode(rx_values), "\n")
```

---

# 15. Scatter Plot

**Age at Start vs. Number of Asthma Prescriptions** — to explore whether older patients at study entry tend to have more asthma prescriptions.

```{r scatter-plot, fig.width=7, fig.height=5}
ggplot(analysis_df_renamed, aes(x = Age_At_Start, y = Asthma_Rx_Count)) +
  geom_point(alpha = 0.3, color = "#2C7BB6", size = 1.2) +
  geom_smooth(method = "lm", color = "#D7191C", se = TRUE) +
  labs(
    title    = "Scatter Plot: Age at Start vs. Number of Asthma Prescriptions",
    subtitle = "Each point represents one patient",
    x        = "Age at Study Start (Years)",
    y        = "Number of Asthma Prescriptions"
  ) +
  theme_minimal(base_size = 12)
```

---

# 16. Bar Plot

**Distribution of Asthma Prescription Counts by Gender** — comparing the number of asthma prescriptions across male and female patients.

```{r bar-plot, fig.width=7, fig.height=5}
bar_data <- analysis_df_renamed %>%
  group_by(Gender) %>%
  summarise(Mean_Asthma_Rx = round(mean(Asthma_Rx_Count), 2))

ggplot(bar_data, aes(x = Gender, y = Mean_Asthma_Rx, fill = Gender)) +
  geom_bar(stat = "identity", width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = Mean_Asthma_Rx), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("#4DAC26", "#D01C8B")) +
  labs(
    title    = "Bar Plot: Mean Asthma Prescriptions by Gender",
    subtitle = "Average number of asthma prescriptions per gender group",
    x        = "Gender",
    y        = "Mean Number of Prescriptions"
  ) +
  theme_minimal(base_size = 12)
```

---

# 17. Pearson Correlation

Finding the correlation between **Age at Start** and **Number of Asthma Prescriptions**.

```{r pearson}
pearson_result <- cor.test(
  analysis_df_renamed$Age_At_Start,
  analysis_df_renamed$Asthma_Rx_Count,
  method = "pearson"
)

cat("Pearson Correlation Coefficient (r):", round(pearson_result$estimate, 4), "\n")
cat("P-value:", format(pearson_result$p.value, scientific = TRUE), "\n")
cat("95% Confidence Interval:", round(pearson_result$conf.int, 4), "\n")

if (abs(pearson_result$estimate) < 0.2) {
  cat("Interpretation: Very weak correlation\n")
} else if (abs(pearson_result$estimate) < 0.4) {
  cat("Interpretation: Weak correlation\n")
} else if (abs(pearson_result$estimate) < 0.6) {
  cat("Interpretation: Moderate correlation\n")
} else {
  cat("Interpretation: Strong correlation\n")
}
```

---

# Conclusion

This analysis explored the Food Allergy dataset from Zenodo, which contains over 333,000 de-identified patient records. Key steps included:

- **Data exploration:** Reviewed structure, variables, and top rows to understand the dataset.
- **Data cleaning:** Removed missing values and duplicate records to ensure data integrity.
- **Feature engineering:** Created new variables such as asthma duration and prescription count doubled.
- **Statistical analysis:** Computed mean, median, mode, and range for key numerical variables.
- **Visualization:** Scatter and bar plots revealed patterns in asthma prescriptions by age and gender.
- **Correlation:** Pearson correlation assessed the linear relationship between age at study entry and asthma prescription frequency.

---

*GitHub Repository: [Add your GitHub link here]*
